Phylogram for testing neutral evolution of a trait, chronogram to model selection.

### 2025-03-17 manuscript/main.tex
Updated manuscript with Théo. G comments

\renewcommand{\baselinestretch}{1.5}
\renewcommand{\arraystretch}{1.2}
\linenumbers
\frenchspacing

@@

% Abstract (≤ 300 words)
\begin{abstract}
The evolution of traits across different species is driven by a combination of selective and neutral processes, while their relative importance is subject to numerous debates. To disentangle them, different scenarios of evolution are modelled and compared. Typically, a neutrally evolving trait is modelled as changing randomly along the different branches of the species tree. This species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time. Indeed, the time that has passed correlates with the changes for a neutral trait, through the direct proxy of the number of generations that occurred. But, we argue that since species generation time can also vary along the phylogenetic tree, chronogram induces biases. Instead, for a phylogram, the branch lengths represent sequence divergence (i.e. the number of substitutions) which absorb the effect of changing generation time. Passed on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a Brownian process has more support on a phylogram than on a chronogram. However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions. Given these results, we argue that we should use phylograms instead of chronograms when claiming that a trait is evolving under drift. Nevertheless, we support the fact that we should generally continue to use chronograms to model selection acting on a trait.
\end{abstract}

@@

%TC:endignore
\section*{Lay summary}\label{sec:summary}
When we look at different species, we can see that some traits are different between them.
For example, the size of the brain or the expression level of a gene varies between species.
The question we ask is: do the time that have passed since the species diverged explain the difference in traits?
Or is it the genetic differences between species that explains best the differences?
The standard in evolutionary biology is to use the time to explain the differences in traits.
However, if the trait is evolving neutrally, the genetic differences between species instead of time should explain best the differences in traits.
As a result, we argue that using simultaneously both the genetic differences and the time since species diverged provides a better understanding of how traits evolve across species.

\section*{Introduction}\label{sec:introduction}
% The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
By observing the changes in trait along a lineage, it is possible to speculate on the regime of selection acting on it.
But without selection, a trait might still changes due to random genetic drift.
As such distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed changes.
At a larger scale, from the observed variations of traits across species, such regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
For example, to model neutral evolution, the mean trait value is said to follow a Brownian Motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
Or in other words, for each  branch of the tree the value at descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
In this framework, reconstructing the trait variation along the whole phylogeny as a BM can thus constitute a null model of neutral trait evolution.

% Alternative models of trait evolution and their pitfalls.
Alternatively to a simple BM, a trend in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, when trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
Methodologically, such alternative models of evolution raise issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a change in the fitness peak~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
Finally, the optimum can also be allowed to change continuously along the tree, which is interpreted as a trait tracking a moving optimum~\citep{hansen_translating_1996, latrille_detecting_2024}.
One special case a continuously changing optimum is again the BM, where the optimum is changing along the phylogenetic tree as a random walk~\citep{hansen_comparative_2008, hansen_three_2024, holstad_evolvability_2024}.
This modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a fluctuating optimum.

% This comes from the fact that we use chronogram, and not phylogram.
% The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
One blind spot of this modelling framework is the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).
Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
Theoretically, for a neutrally evolving trait, trait changes depends directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the generation time is constant.
But, since species generation time might vary along the phylogenetic tree, modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.
Instead, on a phylogram, the branch lengths can represent another quantity than time, which can be used as the backbone to model trait evolution.
For example, it is known that using phylograms in unit of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}.
From a genomic perspective, phylogram can also be obtained in unit of nucleotide divergence, which is equivalent to measuring the number of nucleotide substitutions occurring along a branch.
Such a phylogram in unit of nucleotide divergence, if obtained from neutrally evolving genomic loci, would absorb the effect of changing generation time and mutation rates, and thus would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
Conversely, for a trait evolving under selection, using a chronogram would be more accurate since changes in the optimum fitness peak is extrinsic to the species and thus should be better correlated to time rather than the number of substitutions that occurred along a branch~\citep{hansen_comparative_2008}.
% Finally, from a statistical perspective, using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks) to estimate the branch lengths of chronograms.

@@

% Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
In this study, we test if it is more accurate to use a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram as the underlying tree to model neutral trait changes (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
In other words, we seek to test if change of a neutral trait is better predicted a phylogram.
Conversely, we also seek to test if changes of a trait under a moving optimum is better predicted by a chronogram.
Altogether, by using genomic information we seek to provide additional approaches to model trait evolution.

\section*{Methods}
\subsection*{Simulations along a phylogenetic tree}
We performed simulations under different selective regimes (neutral, moving optimum, multiple optima).
Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree.
We used the same simulation framework as in \citet{latrille_detecting_2024}, with parameters detailed in the supplementary material (section~\ref{sec:simulator}).
The parameters of simulations were chosen to mimic an empirical dataset of mammals.
In summary, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
Parents were selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
Flattening the fitness landscape result in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
Alternatively for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch, if a switch occurs the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).

At each node of the tree, the population is split into two daughter populations running independently on each of the two branches, and the process is repeated until the tips of the tree are reached (Fig.~\ref{fig:results-brownian}C).
As a control, we performed simulations with constant mutation rate, generation times and effective population sizes ($\Ne$).
Alternatively, we performed simulations with fluctuating mutation rate, generation times and $\Ne$, were we used a Brownian process to model the long-term changes along the phylogenetic tree, and we overlaid short-term changes in $\Ne$ (see supplementary material section~\ref{sec:simulator}).

@@

\subsection*{Brownian Motion (BM)}
We used \textit{RevBayes}~\citep{hohna_revbayes_2016} to fit a Brownian Motion (BM) on continuous traits evolution evolving along the branches of a phylogenetic tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a}.
The data consist of the mean trait value for each extant species, and the tree topology is fixed.
Along each branch, the value of the trait is drawn from a normal distribution with mean equal to the parent node value and variance equal to the branch length~\citep{felsenstein_phylogenies_1985}.
We reconstructed the ancestral trait value at each node of the tree as the posterior mean estimate (burn-in of 1000 gen., running of 10000 gen., 2 chains), and we compared the accuracy of the reconstruction using either a phylogram or a chronogram as the underlying tree.
Both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
Importantly, the data and the tree topology are the same in both analyses, only the branch lengths are different.

Additionally to a single rate BM, we fitted a BM with multiple rate parameters~\citep[auteur]{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}), and the number of rate shifts is estimated by reversible-jump MCMC (\textit{dnReversibleJumpMixture} in \textit{RevBayes}).
The number of rate shifts as well as the variance of rate parameters are estimated as the posterior mean (burn-in of 1000 gen., running of 50000 gen., 2 chains).

Finally, we implemented a model in \textit{RevBayes} containing a switch variable, denoted $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
The tree topology is fixed (same for both trees), and both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
As in the previous section, a Brownian process is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length and the squared value of the Brownian rate parameter ($\sigma$, log-uniform prior).
Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favoured over the chronogram.

\subsection*{Ornstein-Uhlenbeck (OU)}
% Simple OU model versus a brownian process.
First, we compared the fit of an Ornstein-Uhlenbeck (OU) process (eq.~\ref{eq:ou-process}) to a BM~\citep{hansen_stabilizing_1997,butler_phylogenetic_2004}, using either a phylogram or a chronogram as the underlying tree.
% Ornstein-Uhlenbeck model equations
\begin{align}
    \der \Trait_t & = -\alpha \left( \Trait_t - \theta \right) \der t + \sigma \der W_t \label{eq:ou-process}
\end{align}
In the OU process, the parameter $\alpha$ in eq.~\ref{eq:ou-process} is the strength of the pull towards the optimum, and $\theta$ is the optimum value.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and because the OU process has more parameters than the Brownian process, we used a reversible-jump MCMC switch between the two models ($0$ for BM and $1$ for OU, \textit{dnReversibleJumpMixture} in \textit{RevBayes}).
% alpha ~ dnReversibleJumpMixture(0.0, dnExponential( abs(root_age / 2.0 / ln(2.0)) ), 0.5)
Mathematically, $\alpha$ is drawn from a reversible-jump mixture distribution: a value of $0$ or drawn from an exponential distribution.
Thus, when $\alpha = 0$ the OU process is equivalent to a BM.
The support for the OU model is estimated as the posterior mean of the switch variable, or equivalently that of $\alpha \neq 0$ (burn-in of 1000 gen., running of 10000 gen., 2 chains).

% Relaxed OU model, how many optimums?
Second, we fitted a relaxed OU with multiple optima~\citep{uyeda_novel_2014}, using either a phylogram or a chronogram as the underlying tree.
The likelihood of the data given the tree is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and the number of optimums is also estimated by the reversible-jump MCMC method (\textit{dnReversibleJumpMixture} in \textit{RevBayes}) as the posterior mean estimate (burn-in of 1000 gen., running of 5000 gen., 2 chains).

@@

\subsection*{Support for a phylogram over a chronogram}
% Simulations of a trait evolving neutrally versus under a moving optimum.
We developed a Bayesian model to estimate the support of phylogram over a chronogram for a trait evolving under a Brownian Motion (BM) process, denoted as $\PhyloSupport$, which range from 0 (full support for the chronogram) to 1.0  (full support for the phylogram), for more details see Methods (BM section).
To test this model, we simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum and 3) under multiple optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
As a control experiment we simulated with a constant generation time, mutation rate (per generation) and effective population size ($\Ne$).
In such a case, the support is the same for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
Next, we simulated changing generation time, mutation rate and $\Ne$ along the phylogenetic tree, mimicking a mammalian range of changes.
In this case, the BM was better fitted on a phylogram than on a chronogram for a neutral trait (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
Conversely, the fit of a BM on a chronogram was better than on a phylogram for a trait under selection, with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).
On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$ when not splitting the males and females.
When splitting the dataset, the support for the phylogram is $\PhyloSupport_{\text{\Female}} = 0.85$ and $\PhyloSupport_{\text{\Male}} = 0.70$.
For the brain mass, the support for the phylogram is $\PhyloSupport = 0.0035$ on the mixed dataset, with $\PhyloSupport_{\text{\Female}} = 0.56$ and $\PhyloSupport_{\text{\Male}} = 0.51$.
On the \textit{COMBINE} dataset, which does not distinguish males and females, the support for the phylogram is $\PhyloSupport = 0.0$ for body mass and $\PhyloSupport = 0.0015$ for brain mass (Table~\ref{table:results-empirical}).
We also computed $\PhyloSupport$ using an approximate likelihood computation (REML), faster the larger trees, and found similar results (Table~\ref{table:results-empirical}).

@@

>>>>

\renewcommand{\baselinestretch}{1.4}
\renewcommand{\arraystretch}{1.2}
\linenumbers
\frenchspacing

@@

% Abstract (≤ 300 words)
\begin{abstract}
At the species level, the evolution of traits is driven by a combination of selective and neutral forces. To disentangle these processes, different scenarios of evolution are modelled and compared. For traits under selection, the species is seen as tracking a trait optimum, and such optimum can change along the species tree. On the other hand, neutral evolution is modelled as the trait changing randomly along the different branches of the species tree. But regardless of the intricacy of modelling trait changes, the species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time. Rationally, the time that has passed correlates with trait changes, even under a neutral regime, through the direct proxy of the number of generations that occurred. However, we argue that since species generation time can also vary along the phylogenetic tree, chronogram induces biases. Instead, for a phylogram, the branch lengths represent sequence divergence (i.e. the number of substitutions) which absorb the effect of changing generation time. Passed on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a random evolution has more support on a phylogram than on a chronogram. However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions. Given these results, we argue that we should use phylograms instead of chronograms when claiming that a trait is evolving under drift. Nevertheless, we support the fact that we should generally continue to use chronograms to model selection acting on a trait.
\end{abstract}

@@

\newpage
%TC:endignore
\section*{Lay summary}\label{sec:summary}
When we look at different species, we can see that some traits are different between them.
For example, the size of the brain or the expression level of a gene varies between species.
The question we ask is: do the time that have passed since the species diverged explain the difference in traits?
Or is it the genetic differences between species that explains best the differences?
The standard in evolutionary biology is to use the time to explain the differences in traits.
However, if the trait is evolving neutrally, the genetic differences between species instead of time should explain best the differences in traits.
As a result, we argue that using simultaneously both the genetic differences and the time since species diverged provides a better understanding of how traits evolve across species.

\section*{Introduction}\label{sec:introduction}
% The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
By observing the changes in trait along a lineage, it is possible to speculate on the regime of selection acting on it.
But even without selection, a trait might still changes due to random genetic drift.
As such distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed changes.
At a larger scale, from the observed variations of traits across species, such regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
For example, to model neutral evolution, the mean trait value is said to follow a Brownian motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
Or in other words, for each  branch of the tree the value at descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
In this framework, reconstructing the trait variation along the whole phylogeny as a BM can thus constitute a null model of neutral trait evolution.

% Alternative models of trait evolution and their pitfalls.
Alternatively to a simple BM, a trend in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, when trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
Methodologically, such alternative models of evolution raise issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a change in the fitness peak~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
Finally, the optimum can also be allowed to change continuously along the tree, which is interpreted as a trait tracking a moving optimum~\citep{hansen_translating_1996, latrille_detecting_2024}.
One special case a continuously changing optimum is again the BM, where the optimum is changing along the phylogenetic tree as a random walk~\citep{hansen_comparative_2008, hansen_three_2024, holstad_evolvability_2024}.
This modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a continuously moving optimum.

% This comes from the fact that we use chronogram, and not phylogram.
% The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
One blind spot of this modelling framework is the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).
Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
Theoretically, for a neutrally evolving trait, trait changes depends directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the generation time is constant.
But, since species generation time might vary along the phylogenetic tree, modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.
Instead, on a phylogram, the branch lengths can represent another quantity than time, which can be used as the backbone to model trait evolution.
For example, it is known that using phylograms in unit of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}.
From a genomic perspective, phylogram can also be obtained in unit of nucleotide divergence, which is equivalent to measuring the number of nucleotide substitutions occurring along a branch.
Such a phylogram in unit of nucleotide divergence, if obtained from neutrally evolving genomic loci, would absorb the effect of changing generation time and mutation rates, and thus would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
Conversely, for a trait evolving under selection at the species level, using a chronogram would be more accurate to model mean trait changes since shifts in the optimum fitness peak is extrinsic to the species, and thus should be better correlated to time rather than the number of substitutions that occurred along a branch~\citep{hansen_comparative_2008}.
% Finally, from a statistical perspective, using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks) to estimate the branch lengths of chronograms.

@@

% Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
In this study, we test if using a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram can allow to assess the regime of evolution of a trait (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
More precisely, we seek to test if change of a neutral trait is better predicted a phylogram.
Conversely, we also seek to test if changes of a trait under a moving optimum is better predicted by a chronogram.
As such, by using genomic information we seek to provide additional approaches to model trait evolution, particularly to disentangle selection and neutral evolution.

\section*{Methods}
In order to assess whether a phylogram is better suited to model neutral evolution we first performed simulations of a trait evolving along a phylogenetic tree under different regimes of selection (neutral, moving optimum, multiple optima).
Second, we gathered an empirical dataset of body and brain masses from mammals, including both a chronogram and a phylogram (in unit of nucleotide divergence from neutral loci).
Third, we fitted a single rate Brownian motion (BM) to the simulated data, and we compared the fit of the BM on a phylogram versus a chronogram.
Forth, based in the single rate BM, we derived a model to estimate the support of a phylogram over a chronogram that we apply to the simulated and empirical datasets.
Finally, we fitted several models of trait evolution to the simulated and empirical datasets: a multi-rate BM, an Ornstein-Uhlenbeck (OU) process, and a relaxed OU with multiple optima.

\subsection*{Simulations along a phylogenetic tree}
We performed simulations under different selective regimes (neutral, moving optimum, multiple optima).
Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree.
We used the same simulation framework as in \citet{latrille_detecting_2024}, with parameters detailed in the supplementary material (section~\ref{sec:simulator}).
The parameters of simulations were chosen to mimic an empirical dataset of mammals.
In summary, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
Parents were selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
Flattening the fitness landscape result in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
Alternatively for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing as a geometric Brownian motion (BM) along the phylogenetic tree~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch, if a switch occurs the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).

At each node of the tree, the population is split into two daughter populations running independently on each of the two branches, and the process is repeated until the tips of the tree are reached (Fig.~\ref{fig:results-brownian}C).
As a control, we performed simulations with constant mutation rate, generation times and effective population sizes ($\Ne$).
Alternatively, we performed simulations with fluctuating mutation rate, generation times and $\Ne$, were we used a BM to model the long-term changes along the phylogenetic tree, and we overlaid short-term changes in $\Ne$ (see supplementary material section~\ref{sec:simulator}).

@@

\subsection*{Brownian motion (BM)}
On the simulated dataset, we fitted a single rate BM using either a phylogram or a chronogram as the underlying tree.
We used \textit{RevBayes}~\citep{hohna_revbayes_2016} to fit a Brownian motion (BM) on continuous traits evolution evolving along the branches of a phylogenetic tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a}.
The data consist of the mean trait value for each extant species, and the tree topology is fixed.
Along each branch, the value of the trait is drawn from a normal distribution with mean equal to the parent node value and variance equal to the branch length~\citep{felsenstein_phylogenies_1985}.
We reconstructed the ancestral trait value at each node of the tree as the posterior mean estimate (burn-in of 1000 gen., running of 10000 gen., 2 chains), and we compared the accuracy of the reconstruction using either a phylogram or a chronogram as the underlying tree.
Both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
Importantly, the data and the tree topology are the same in both analyses, only the branch lengths are different.
We expect for trait under neutral evolution more accurate trait reconstruction on a phylogram, and conversely for trait under selection (moving optimum) more accurate trait reconstruction on a chronogram.

\subsection*{BM with a switch}
To test the support of a phylogram over a chronogram for the simulated and empirical datasets, we implemented a model based on a single rate BM.
Such model was implemented in \textit{RevBayes} and contains a switch variable, denoted $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
The tree topology is fixed (same for both trees), and both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
As in the previous section, a BM is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length and the squared value of the Brownian rate parameter ($\sigma$, log-uniform prior).
Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favoured over the chronogram.
We expect $\PhyloSupport = 1$ for trait under neutral evolution, and $\PhyloSupport = 0$ for trait under selection (moving optimum).

\subsection*{Alternatives to single rate BM}
Additionally to a single rate BM, we fitted a BM with multiple rate parameters~\citep[auteur]{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree to the simulated dataset.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}), and the number of rate shifts is estimated by reversible-jump MCMC (\textit{dnReversibleJumpMixture} in \textit{RevBayes}).
The number of rate shifts as well as the variance of rate parameters are estimated as the posterior mean (burn-in of 1000 gen., running of 50000 gen., 2 chains).

% Simple OU model versus a brownian process.
Also, we compared the fit of an Ornstein-Uhlenbeck (OU) process (eq.~\ref{eq:ou-process}) to a BM~\citep{hansen_stabilizing_1997,butler_phylogenetic_2004}, using either a phylogram or a chronogram as the underlying tree.
% Ornstein-Uhlenbeck model equations
\begin{align}
    \der \Trait_t & = -\alpha \left( \Trait_t - \theta \right) \der t + \sigma \der W_t \label{eq:ou-process}
\end{align}
In the OU process, the parameter $\alpha$ in eq.~\ref{eq:ou-process} is the strength of the pull towards the optimum, and $\theta$ is the optimum value.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and because the OU process has more parameters than the BM, we used a reversible-jump MCMC switch between the two models ($0$ for BM and $1$ for OU, \textit{dnReversibleJumpMixture} in \textit{RevBayes}).
% alpha ~ dnReversibleJumpMixture(0.0, dnExponential( abs(root_age / 2.0 / ln(2.0)) ), 0.5)
Mathematically, $\alpha$ is drawn from a reversible-jump mixture distribution: a value of $0$ or drawn from an exponential distribution.
Thus, when $\alpha = 0$ the OU process is equivalent to a BM.
The support for the OU model is estimated as the posterior mean of the switch variable ($p_{\text{OU}}$), or equivalently that of $\alpha \neq 0$ (burn-in of 1000 gen., running of 10000 gen., 2 chains).

% Relaxed OU model, how many optimums?
Finally, we fitted a relaxed OU with multiple optima~\citep{uyeda_novel_2014}, using either a phylogram or a chronogram as the underlying tree.
The likelihood of the data given the tree is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and the number of optimums is also estimated by the reversible-jump MCMC method (\textit{dnReversibleJumpMixture} in \textit{RevBayes}) as the posterior mean estimate (burn-in of 1000 gen., running of 5000 gen., 2 chains).

@@

\subsection*{Support for a phylogram over a chronogram}
% Simulations of a trait evolving neutrally versus under a moving optimum.
We first tested the hypothesis that a phylogram is better suited to model neutral evolution than a chronogram.
We simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum and 3) under multiple optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
We tested for the support of phylogram over a chronogram under the assumption that changes follow a Brownian motion (BM) process, we developed as statistic denoted as $\PhyloSupport$, which range from 0 (full support for the chronogram) to 1.0 (full support for the phylogram).
For simulations with a constant generation time, the support is similar for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
Next, we simulated changing generation time, mutation rate (per generation) and effective population size ($\Ne$) along the phylogenetic tree, mimicking a mammalian range of changes.
In this case, for a neutral trait, the BM was better fitted on a phylogram than on a chronogram (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
Conversely, for a trait under selection, the fit of a BM on a chronogram was better than on a phylogram with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).

On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$ when not splitting the males and females.
When splitting the dataset, the support for the phylogram is $\PhyloSupport_{\text{\Female}} = 0.85$ and $\PhyloSupport_{\text{\Male}} = 0.70$.
For the brain mass, the support for the phylogram is $\PhyloSupport = 0.0035$ on the mixed dataset, with $\PhyloSupport_{\text{\Female}} = 0.56$ and $\PhyloSupport_{\text{\Male}} = 0.51$.
On the \textit{COMBINE} dataset, which does not distinguish males and females, the support for the phylogram is $\PhyloSupport = 0.0$ for body mass and $\PhyloSupport = 0.0015$ for brain mass (Table~\ref{table:results-empirical}).
We also computed $\PhyloSupport$ using an approximate likelihood computation (REML), faster the larger trees, and found similar results (Table~\ref{table:results-empirical}).

@@

\printbibliography
